---
title: "Random Case Effect on Reading and Other Reading Phenomenon"
author: Tyler Rinker
output:
  md_document:
    toc: true
    toc_depth: 2  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## rmarkdown::render("README.Rmd", "all")
```

This repo is a demonstration of the random mixed case effect and other reading phenomenon on the ability to read.   

# Random Case Effect

[Mayall, Humphreys & Olson (1997)](articles/Mayall1997.pdf) document this effect.

> Mayall, K., Humphreys, G. W., & Olson, A. (1997). Disruption to word or letter processing?: The origins of case-mixing effects. Journal of Experimental Psychology: Learning, Memory, and Cognition, 23(5), 1275-1286. 10.1037/0278-7393.23.5.1275 (on UBlearns)


The authors argue that mixing case inappropriately has a negative impact on the ability for readers to decode text.  I wanted to test this effect on myself.  I needed a sentence and decided to make a minimal R function to randomly alter case:

```{r}
random_upper <- function(x, prop = .5, ...){

    stopifnot(prop > 0 & prop <= 1)

    y <- strsplit(x, '')

    unlist(lapply(y, function(a){
        locs <- grepl('[a-z]', a)   
        ilocs <- which(locs)  
        to_upper <- sample(ilocs, ceiling(prop * length(ilocs)))  
        a[to_upper] <- toupper(a[to_upper])
        paste(a, collapse = '')
    }))
}

x <- "Many English words are formed by taking basic words and adding combinations of prefixes and suffixes to them."
```


```{r}
random_upper(x, .1)
random_upper(x, .3)
random_upper(x, .5)
```

Indeed, I informally recognize that mixing case made decoding harder (likely removing automaticity in sight word recognition).

# Random Scramble

The effect of scrambling internal letters (not first or last letter) began as a semi-true Internet meme hoax.  The hoax read as follows:

> It deosn't mttaer in waht oredr the ltteers in a wrod are, the olny iprmoetnt tihng is taht the frist and lsat ltteer be at the rghit pclae. The rset can be a toatl mses and you can sitll raed it wouthit porbelm. Tihs is bcuseae the huamn mnid deos not raed ervey lteter by istlef, but the wrod as a wlohe.
 
Obviously, there is an element of truth to this but indeed there is a cost to scrambling letters.  [Rayner, White, Johnson, & Liversedge (2006)](articles/Mayall1997.pdf) document this effect.

> Rayner, K., White, S. J., Johnson, R. L., & Liversedge, S. P. (2006). Raeding wrods with jubmled lettres: There is a cost. Psychological Science, 17(3), 192-193. 10.1111/j.1467-9280.2006.01684.x 

The code below produces random resamples of internal letters.  The `window` argument controls the window function (what letters are kept together), while the `sample.grams` argument is a logical command that, if true, reorders the gram sequences that have been sampled.  For example, let's say I had the sequence `123456`.  Sampling grams of length 3 may produce `231564`.  Setting ` sample.grams = TRUE` may further produce `564231`.


```{r}
random_scramble <- function(x, window = 2, sample.grams = TRUE, ...){

    if (!is.integer(window)) window <- as.integer(window)
    stopifnot(all(window > 1))

    y <- textshape::split_token(x, lower = FALSE)

    unlist(lapply(y, function(a){
        out <- unlist(lapply(a, function(b){

            ## get a sample from window if not fixed
            if (length(window) > 1) w <-  sample(window, 1) else w <- window

            ## ensure length of word is > 2
            len <- nchar(b)
            if (len < 4) return(b)

            z <- strsplit(b, '')[[1]]
            locs <- 2:(len - 1)

            if (length(locs) <= w) {
                return(paste(c(z[1], paste(sample(z[locs]), collapse = ''), z[length(z)]), collapse = ''))
            }

            locs2 <- rep(1:floor(length(locs)/w), each = w)

            locs3 <- rle(c(locs2, c(rep(max(locs2) + 1, length(locs)%%w))))

            locs4 <- rep(locs3$values, sample(locs3$lengths))

            rands <- unlist(lapply(split(locs, locs4), function(grams){
                paste(sample(z[grams]), collapse = '')
            }))

            if (sample.grams) rands <- sample(rands)

            paste(c(z[1], rands, z[length(z)]), collapse = '')  

        }))

        trimws(gsub("(\\s+)([.!?,;:])", "\\2", paste(out, collapse = ' '), perl = TRUE))

    }))

}



x <- c(
    "According to a study at an English University, it doesn't matter in what order the letters in a word are, the only important thing is that the first and last letter be at the right place. The rest can be a total mess and you can still read it without problem. This is because the human mind does not read every letter by itself but the word as a whole.",
    "A vehicle exploded at a police checkpoint near the UN headquarters in Baghdad on Monday killing the bomber and an Iraqi police officer",
    "Big council tax increases this year have squeezed the incomes of many pensioners",
    "A doctor has admitted the manslaughter of a teenage cancer patient who died after a hospital drug blunder."
)
```

```{r}
random_scramble(x, 2)
random_scramble(x, 2, sample.grams = FALSE)
random_scramble(x, 2:5)
random_scramble(x, 5)
```
